---
title: "Optimal Fire Station Placement on Oahu"
subtitle: "A Spatial Optimization Analysis"
format: html
editor: visual
---

## Introduction

This analysis identifies the optimal location for a **new fire station on Oahu** to maximize emergency response coverage. Using spatial optimization, we combine population data, current coverage gaps, and land use suitability to produce a data-driven recommendation.

**Research Question:** Where are the spatial gaps in fire station coverage on Oahu, and where should a new fire station be located to maximize emergency response accessibility while respecting land use constraints?

------------------------------------------------------------------------

## Setup

```{r}
#| message: false
#| warning: false

library(tidyverse)
library(tidycensus)
library(sf)
library(mapview)
library(spopt)
library(osmdata)
library(RColorBrewer)
```

------------------------------------------------------------------------

## Step 1: Load and Filter Data to Main Island

We load fire stations and census tracts, then immediately project to UTM Zone 4N (EPSG:32604) for accurate distance calculations in meters. Because Honolulu County includes the Northwestern Hawaiian Islands, we filter to the main Oahu island only using easting coordinates.

```{r}
#| message: false
#| warning: false

# Load existing fire stations
fire_stations_raw <- st_read(
  "/Users/berylinlau/Downloads/Fire_Stations_-_Island_of_Oahu.geojson",
  quiet = TRUE
)

# Load population data for Honolulu County
oahu_tracts_raw <- get_acs(
  geography = "tract",
  variables = "B01003_001",
  state = "HI",
  county = "Honolulu",
  geometry = TRUE,
  year = 2022
)

# Project both to UTM Zone 4N immediately
fire_stations_raw <- st_transform(fire_stations_raw, 32604)
oahu_tracts_raw   <- st_transform(oahu_tracts_raw, 32604)

# Filter to main Oahu island only (easting > 500,000 m removes outer islands)
oahu_tracts <- oahu_tracts_raw |>
  mutate(centroid_x = st_coordinates(st_centroid(geometry))[, 1]) |>
  filter(centroid_x > 500000) |>
  select(-centroid_x)

oahu_boundary <- st_union(oahu_tracts) |> st_make_valid()

fire_stations <- fire_stations_raw |>
  filter(lengths(st_intersects(fire_stations_raw, oahu_boundary)) > 0)

cat("Census tracts (main island):", nrow(oahu_tracts), "\n")
cat("Fire stations (main island):", nrow(fire_stations), "\n")
cat("Total population:", format(sum(oahu_tracts$estimate), big.mark = ","), "\n")

# Compute bbox in lon/lat for OSM downloads
oahu_bbox <- st_bbox(st_transform(oahu_tracts, 4326))
```

```{r}
# Map existing fire stations
mapview(oahu_tracts,
        layer.name = "Census Tracts",
        alpha.regions = 0.1) +
  mapview(fire_stations,
          col.regions = "red",
          cex = 6,
          layer.name = "Existing Fire Stations")
```

Oahu's main island has **`r nrow(fire_stations)` fire stations** serving a population of **`r format(sum(oahu_tracts$estimate), big.mark = ",")`** across **`r nrow(oahu_tracts)` census tracts**.

------------------------------------------------------------------------

## Step 2: Download and Classify Land Use

We download land use polygons from OpenStreetMap and assign each a **suitability score (1--5)** for fire station placement, based on how easily the land could be converted or acquired.

```{r}
#| message: false
#| warning: false

sf_use_s2(FALSE)

# Download land use polygons
landuse_raw <- opq(bbox = oahu_bbox) |>
  add_osm_feature(key = "landuse") |>
  osmdata_sf()

landuse_classified <- landuse_raw$osm_polygons |>
  select(osm_id, landuse) |>
  st_transform(32604) |>
  st_make_valid() |>
  mutate(
    zone_type = case_when(
      landuse %in% c("residential", "apartments")                           ~ "Residential",
      landuse %in% c("commercial", "retail")                                ~ "Commercial",
      landuse %in% c("industrial", "construction")                          ~ "Industrial",
      landuse %in% c("military", "government")                              ~ "Government/Public",
      landuse %in% c("recreation_ground", "park", "grass",
                     "cemetery", "village_green")                           ~ "Open Space/Parks",
      landuse %in% c("institutional", "education", "hospital", "religious") ~ "Institutional",
      landuse %in% c("farmland", "farm", "orchard", "meadow")              ~ "Agricultural",
      landuse == "forest"                                                    ~ "Forest",
      TRUE                                                                   ~ "Other"
    ),
    suitability = case_when(
      zone_type == "Government/Public"                      ~ 5,
      zone_type %in% c("Open Space/Parks", "Commercial")   ~ 4,
      zone_type %in% c("Institutional", "Industrial")      ~ 3,
      zone_type %in% c("Residential", "Agricultural")      ~ 2,
      TRUE                                                  ~ 1
    ),
    conversion_category = case_when(
      suitability >= 4 ~ "No Conversion Needed",
      suitability == 3 ~ "Minor Rezoning Required",
      suitability == 2 ~ "Major Conversion Required",
      TRUE             ~ "Not Feasible"
    )
  )

cat("Land use polygons downloaded:", nrow(landuse_classified), "\n")
```

```{r}
mapview(landuse_classified,
        zcol = "suitability",
        layer.name = "Land Use Suitability (1-5)",
        col.regions = RColorBrewer::brewer.pal(5, "RdYlGn"),
        alpha.regions = 0.7) +
  mapview(fire_stations,
          col.regions = "blue",
          cex = 7,
          layer.name = "Existing Stations")
```

**Suitability scale:** 5 = Government/Public land (ideal) → 1 = Forest (not feasible)

------------------------------------------------------------------------

## Step 3: Calculate Current Coverage

We calculate the straight-line distance from each census tract centroid to the nearest fire station, with a 1.5× terrain adjustment factor to approximate road travel distance. Tracts are classified into three coverage tiers.

```{r}
tract_centroids <- st_centroid(oahu_tracts)
dist_matrix     <- st_distance(tract_centroids, fire_stations)
dist_matrix_adj <- as.matrix(dist_matrix) * 1.5

oahu_tracts <- oahu_tracts |>
  mutate(
    dist_to_nearest_m     = apply(dist_matrix_adj, 1, min),
    dist_to_nearest_miles = dist_to_nearest_m / 1609.34,
    coverage_status = case_when(
      dist_to_nearest_miles <= 2 ~ "Well Covered (< 2 mi)",
      dist_to_nearest_miles <= 4 ~ "Moderate Coverage (2-4 mi)",
      TRUE                       ~ "Poor Coverage (> 4 mi)"
    )
  )

oahu_tracts$coverage_status <- factor(
  oahu_tracts$coverage_status,
  levels = c("Well Covered (< 2 mi)", "Moderate Coverage (2-4 mi)", "Poor Coverage (> 4 mi)")
)

coverage_summary <- oahu_tracts |>
  st_drop_geometry() |>
  group_by(coverage_status) |>
  summarize(
    n_tracts        = n(),
    total_pop       = sum(estimate),
    pct_population  = round(total_pop / sum(oahu_tracts$estimate) * 100, 1),
    mean_dist_miles = round(mean(dist_to_nearest_miles), 1),
    .groups = "drop"
  ) |>
  arrange(coverage_status)

knitr::kable(coverage_summary,
             col.names = c("Coverage Status", "Tracts", "Population",
                           "% of Population", "Mean Distance (mi)"),
             caption = "Current fire station coverage on Oahu")
```

```{r}
mapview(oahu_tracts,
        zcol = "coverage_status",
        layer.name = "Coverage Status",
        col.regions = c("lightgreen", "yellow", "red"),
        alpha.regions = 0.65) +
  mapview(fire_stations,
          col.regions = "blue",
          cex = 7,
          layer.name = "Existing Stations")
```

**`r coverage_summary$pct_population[coverage_summary$coverage_status == "Poor Coverage (> 4 mi)"]`%** of Oahu's population lives more than 4 miles from the nearest fire station.

------------------------------------------------------------------------

## Step 4: Calculate Response Need Score

We create a composite **Total Need Score** combining two factors:

-   **Distance (60% weight):** Tracts farther from stations score higher
-   **Population density (40% weight):** Denser tracts score higher

This ensures the optimization prioritizes areas that are both underserved *and* populous.

```{r}
oahu_tracts <- oahu_tracts |>
  mutate(
    area_sq_miles      = as.numeric(st_area(geometry)) / 2589988.11,
    pop_density        = estimate / area_sq_miles,
    distance_scaled    = (dist_to_nearest_miles - min(dist_to_nearest_miles, na.rm = TRUE)) /
                         (max(dist_to_nearest_miles, na.rm = TRUE) - min(dist_to_nearest_miles, na.rm = TRUE)),
    pop_density_scaled = (pop_density - min(pop_density, na.rm = TRUE)) /
                         (max(pop_density, na.rm = TRUE) - min(pop_density, na.rm = TRUE)),
    per_capita_need    = distance_scaled * 60 + pop_density_scaled * 40,
    total_need         = per_capita_need * estimate
  )

oahu_tracts |>
  st_drop_geometry() |>
  arrange(desc(total_need)) |>
  select(NAME, estimate, dist_to_nearest_miles, pop_density, per_capita_need, total_need) |>
  head(10) |>
  knitr::kable(digits = 1, caption = "Top 10 highest-need census tracts")
```

```{r}
mapview(oahu_tracts,
        zcol = "total_need",
        layer.name = "Total Response Need Score",
        col.regions = RColorBrewer::brewer.pal(9, "YlOrRd"),
        alpha.regions = 0.7) +
  mapview(fire_stations,
          col.regions = "blue",
          cex = 6,
          layer.name = "Existing Stations")
```

------------------------------------------------------------------------

## Step 5: Generate Candidate Sites

We create a 20×20 grid of candidate locations across the main island, then filter to points on land that are more than 1 mile from an existing station. Each candidate is assigned a land use suitability score.

```{r}
set.seed(2025)
sf_use_s2(FALSE)

bbox  <- st_bbox(oahu_tracts)
x_seq <- seq(bbox["xmin"], bbox["xmax"], length.out = 20)
y_seq <- seq(bbox["ymin"], bbox["ymax"], length.out = 20)

grid_points <- expand.grid(x = x_seq, y = y_seq) |>
  st_as_sf(coords = c("x", "y"), crs = 32604)

on_land        <- lengths(st_intersects(grid_points, oahu_boundary)) > 0
candidate_grid <- grid_points[on_land, ] |>
  mutate(candidate_id = row_number())

dist_to_existing  <- st_distance(candidate_grid, fire_stations)
min_dist_existing <- apply(dist_to_existing, 1, min)

candidate_grid <- candidate_grid |>
  mutate(min_dist_miles = as.numeric(min_dist_existing) / 1609.34) |>
  filter(min_dist_miles > 1)

candidate_sites <- candidate_grid |>
  st_join(st_make_valid(landuse_classified), join = st_intersects, left = TRUE) |>
  mutate(
    suitability = if_else(is.na(suitability), 3L, as.integer(suitability)),
    zone_type   = if_else(is.na(zone_type), "Unknown/Undeveloped", zone_type),
    conversion_category = case_when(
      suitability >= 4 ~ "No Conversion Needed",
      suitability == 3 ~ "Minor Rezoning Required",
      suitability == 2 ~ "Major Conversion Required",
      TRUE             ~ "Not Feasible"
    )
  ) |>
  group_by(candidate_id) |>
  slice_max(suitability, n = 1, with_ties = FALSE) |>
  ungroup()

candidate_sites_suitable <- candidate_sites |> filter(suitability >= 3)

cat("Total candidates:", nrow(candidate_sites), "\n")
cat("Suitable candidates (suitability >= 3):", nrow(candidate_sites_suitable), "\n")
```

```{r}
mapview(oahu_tracts,
        zcol = "total_need",
        layer.name = "Total Need",
        col.regions = RColorBrewer::brewer.pal(9, "YlOrRd"),
        alpha.regions = 0.4) +
  mapview(candidate_sites,
          zcol = "suitability",
          layer.name = "Candidate Sites (Suitability)",
          col.regions = RColorBrewer::brewer.pal(5, "RdYlGn"),
          cex = 4) +
  mapview(fire_stations,
          col.regions = "red",
          cex = 7,
          layer.name = "Existing Stations")
```

------------------------------------------------------------------------

## Step 6: Spatial Optimization

We use the **p-median algorithm** to find candidate locations that minimize population-weighted distance. We run three scenarios: unconstrained (any land), constrained (suitable land only), and a 3-station constrained solution.

```{r}
if (nrow(candidate_sites_suitable) < 3) {
  candidate_sites_suitable <- candidate_sites
}

result_unconstrained <- p_median(
  demand       = oahu_tracts,
  facilities   = candidate_sites,
  n_facilities = 1,
  weight_col   = "total_need"
)

result_constrained <- p_median(
  demand       = oahu_tracts,
  facilities   = candidate_sites_suitable,
  n_facilities = 1,
  weight_col   = "total_need"
)

result_3stations <- p_median(
  demand       = oahu_tracts,
  facilities   = candidate_sites_suitable,
  n_facilities = 3,
  weight_col   = "total_need"
)

selected_unconstrained <- result_unconstrained$facilities |> filter(.selected)
selected_constrained   <- result_constrained$facilities   |> filter(.selected)
selected_3stations     <- result_3stations$facilities     |> filter(.selected)

cat("Unconstrained optimum zone:", selected_unconstrained$zone_type, "\n")
cat("Constrained optimum zone:  ", selected_constrained$zone_type, "\n")
```

------------------------------------------------------------------------

## Step 7: Results

### Recommended Single Station (Zoning-Constrained)

```{r}
optimal_site <- selected_constrained |> st_drop_geometry()

cat("Zone Type:        ", optimal_site$zone_type, "\n")
cat("Suitability:      ", optimal_site$suitability, "/ 5\n")
cat("Conversion:       ", optimal_site$conversion_category, "\n")
cat("Dist from nearest:", round(optimal_site$min_dist_miles, 1), "miles\n")
```

```{r}
all_stations_1 <- bind_rows(
  fire_stations        |> mutate(type = "Existing"),
  selected_constrained |> mutate(type = "Proposed")
)

dist_new_adj <- as.matrix(st_distance(tract_centroids, all_stations_1)) * 1.5

oahu_tracts_new <- oahu_tracts |>
  mutate(
    dist_new_miles    = apply(dist_new_adj, 1, min) / 1609.34,
    improvement_miles = dist_to_nearest_miles - dist_new_miles
  )

tibble(
  Metric = c("Mean distance to nearest station (miles)",
             "Population within 2 miles",
             "Population beyond 4 miles"),
  Before = c(
    round(mean(oahu_tracts$dist_to_nearest_miles), 2),
    sum(oahu_tracts$estimate[oahu_tracts$dist_to_nearest_miles <= 2]),
    sum(oahu_tracts$estimate[oahu_tracts$dist_to_nearest_miles > 4])
  ),
  After = c(
    round(mean(oahu_tracts_new$dist_new_miles), 2),
    sum(oahu_tracts_new$estimate[oahu_tracts_new$dist_new_miles <= 2]),
    sum(oahu_tracts_new$estimate[oahu_tracts_new$dist_new_miles > 4])
  )
) |>
  mutate(Improvement = Before - After,
         `% Change`  = round((Improvement / Before) * 100, 1)) |>
  knitr::kable(caption = "Coverage improvement: 1 new station",
               format.args = list(big.mark = ","))
```

```{r}
mapview(oahu_tracts_new,
        zcol = "improvement_miles",
        layer.name = "Distance Improvement (miles)",
        col.regions = RColorBrewer::brewer.pal(9, "Blues"),
        alpha.regions = 0.65) +
  mapview(fire_stations,
          col.regions = "red",
          cex = 7,
          layer.name = "Existing Stations") +
  mapview(selected_constrained,
          col.regions = "darkgreen",
          cex = 14,
          layer.name = "Recommended New Station")
```

### 3-Station Solution

```{r}
all_stations_3 <- bind_rows(
  fire_stations      |> mutate(type = "Existing"),
  selected_3stations |> mutate(type = "Proposed")
)

dist_new3_adj <- as.matrix(st_distance(tract_centroids, all_stations_3)) * 1.5

oahu_tracts_new3 <- oahu_tracts |>
  mutate(
    dist_new_miles    = apply(dist_new3_adj, 1, min) / 1609.34,
    improvement_miles = dist_to_nearest_miles - dist_new_miles
  )

tibble(
  Metric = c("Mean distance to nearest station (miles)",
             "Population within 2 miles",
             "Population beyond 4 miles"),
  Before = c(
    round(mean(oahu_tracts$dist_to_nearest_miles), 2),
    sum(oahu_tracts$estimate[oahu_tracts$dist_to_nearest_miles <= 2]),
    sum(oahu_tracts$estimate[oahu_tracts$dist_to_nearest_miles > 4])
  ),
  After = c(
    round(mean(oahu_tracts_new3$dist_new_miles), 2),
    sum(oahu_tracts_new3$estimate[oahu_tracts_new3$dist_new_miles <= 2]),
    sum(oahu_tracts_new3$estimate[oahu_tracts_new3$dist_new_miles > 4])
  )
) |>
  mutate(Improvement = After - Before,
         `% Change`  = round((Improvement / Before) * 100, 1)) |>
  knitr::kable(caption = "Coverage improvement: 3 new stations",
               format.args = list(big.mark = ","))
```

```{r}
mapview(oahu_tracts_new3,
        zcol = "improvement_miles",
        layer.name = "Distance Improvement (miles)",
        col.regions = RColorBrewer::brewer.pal(9, "Blues"),
        alpha.regions = 0.65) +
  mapview(fire_stations,
          col.regions = "red",
          cex = 7,
          layer.name = "Existing Stations") +
  mapview(selected_3stations,
          col.regions = "darkgreen",
          cex = 14,
          layer.name = "3 Recommended New Stations")
```

### Constrained vs. Unconstrained Comparison

```{r}
dist_between <- as.numeric(st_distance(selected_unconstrained, selected_constrained)) / 1609.34

mapview(oahu_tracts,
        zcol = "total_need",
        layer.name = "Total Need",
        col.regions = RColorBrewer::brewer.pal(9, "YlOrRd"),
        alpha.regions = 0.4) +
  mapview(selected_unconstrained,
          col.regions = "orange",
          cex = 14,
          layer.name = "Unconstrained Optimum") +
  mapview(selected_constrained,
          col.regions = "darkgreen",
          cex = 14,
          layer.name = "Feasible Optimum (Zoning-Constrained)") +
  mapview(fire_stations,
          col.regions = "red",
          cex = 7,
          layer.name = "Existing Stations")
```

The unconstrained and zoning-constrained optima are **`r round(dist_between, 2)` miles apart**.

------------------------------------------------------------------------

## Conclusion

This analysis recommends placing a new fire station on **`r selected_constrained$zone_type`** land in the `r if_else(st_coordinates(selected_constrained)[1] < 607000, "western/central", "eastern")` part of Oahu. This location:

-   Addresses the **largest coverage gap** on the island
-   Serves the **highest-need communities** based on population and distance
-   Is situated on land **`r tolower(selected_constrained$conversion_category)`**, making implementation feasible

### Limitations

-   Distances use a terrain-adjusted straight-line approximation rather than true road network routing
-   OSM land use data may be incomplete; official Honolulu zoning data would improve accuracy
-   Analysis does not account for land acquisition costs, ownership, or environmental regulations
-   Future work could incorporate historical call data and traffic patterns
